Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015

### 2017-10-10 paper/paper.Rnw
Rewrote hypothesis section + introduction

\title{Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}

@@

% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}. But fast-forward until today, we've understood the relationship much better.

@@

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills. Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}. Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistance, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.

@@

\section{Research framework}

\subsection{Research questions and hypothesis}

The overaching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.

We develop the subquestions and their corresponding hypothesis separately in more detail.

@@

\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though. After recording a SES gap increase of about 40\% in 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap. First, we will concentrate on the evolution of the gap only for 15 year olds (which as we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap), which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}. Secondly, we will compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature. \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}. We posit no specific hypothesis for this question given that it's purely explanatory.

\begin{enumerate}
\item Research question:
\begin{enumerate}
\item Which countries are experiencing changes in the achievement gap?
\end{enumerate}
\end{enumerate}
\end{enumerate}

b)

The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. Following that strategy will put us closer to understanding the source of the evolution of the achievement gap.

Research question:
* Is the gap originating because the top is gaining ground, the top is falling behind or because of a dyanmic interaction between the two?

Hypothesis 1:
* The theorical arugment in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups. We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-ses groups at a greater disadvantage. Conversely, in countries with low tracking we should expect for the achievement gap to be narrowing.

c)

Our study itself is composed of several cross-sectional surveys, which is limited to single cross-sectional snapshots, but given we have 6 times points for most countries we can model the evolution of the achievement gap over time, something not done before in an international context. We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking}.

Research question:
* Are tracking and vocational enrollment related to the evolution of the gap?

Hypothesis 2:
* Building on the previous hypotheiss, tracking should play an important role on the evolution of the achievement gap. We hypothesize that the degree of tracking of an educational system is tightly related to changes in the gap, and the more tracking, the more inequality. Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.

d)

We want to investigate whether better performing countries have lower levels of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high levels of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

Research question:
* Are achievement and inequality correlated? and if so, has the relationship weakeaned/strengthend over time?

Hypothesis 3:
* Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality. Countries with high performance should have less inequality and countries with low performance should have greater inequality

@@

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

>>>>

\title{Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}

@@

The impact of inequality, be it economic or of non pecuniary causes, is a crucial determinant for the opportunities of a child. Evidence from disciplines as disparate as epidemiology, psychology, sociology and economics keep piling evidence that suggests that inequality in the life course is something present in virtually all societies. This is something that politicians and policymakers are aware of and, yet despite their many efforts, inequality has not abated.

The literature has found conclusive evidence showing that, for example, the United States has increased both economic inequality \citep{piketty2003}, inequality of opportunity \citep{chetty2014} and inequality of achievement \citep{reardon2011}. All of these findings, however, are from the United States. Other countries such as the United Kingdom or other developed countries have their fair share of studies when there is data available. However, very rarely can we compare levels of inequality (of any kind), across several countries with a great degree of confidence.

This paper looks to study the cognitive achievement gap between the 90th and 10th percentile of the SES distribution. This gap can be termed as the SES achievement gap between the top and bottom SES groups. Using the PISA database, we calculate the achievement gap for every country and for all six available PISA waves, building a country panel that shows a time trend of 15 years for each country.

More concretely, we aim to study the evolution of the achievement gap. We aim to answer two broad question: Which countries are changing over time while others are stagnated? Equally important, what explains these changes? Building on the literature on educational institutions we suggest the degree of tracking of a country and it's level of vocational enrollments as possible explanations for the changes of a country over time.

Findings suggest that there's reasonable variation in the evolution of the gap. For example, the United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%. This reduction is also present in other countries such as Germany and Poland. Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap. Robustness checks shows that also the 80/20 and 70/30 replicate the results for nearly all countries.

We then assess whether this gap is correlated with the country's average performance, and find a very strong negative correlation for all country pooled. This suggests that countries with high levels of achievement have low levels of inequality. Yet these results are aggregated for every year. A more disaggregate analysis shows that this pattern is also present for every country over time, although some exceptions are encountered, like the United States where achievement and inequality as positively correlated.

To finish, we find that the degree of tracking of a country seems to be a likely explanation to the changes in the achievement gap, but it should not be studied without factoring in the degree of vocational enrollment. We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap. Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly. Simulations show that if every country switched to little or no tracking their gaps would shrink by about 20\%.

These findings have two specific implications. First, inequality should not be 'generalized' or 'summarized' with one single indicator for many countries. There are widespread differences between countries, where each one should be studied in context. Secondly, now that we know that, further research should attempt to explain why some countries follow such different evolutions. We propose a possible explanation, the degree of tracking and vocational enrollment of the country. Tracking seems to exacerbate inequality, but if coped with a reasonable level of vocational enrollment it can alleviate the negative effects of tracking. Further research should confirm these results hold for other empirical tests and if so, policymakers should incorporate vocational enrollment not only as a 'labor market' solution to youth unemployment but as a compensatory mechanism for highly tracked curriculums.

\section{Literature Review}
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}. But fast-forward until today, we've understood the relationship much better.

@@

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills. Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}. Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistence, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.

@@

\section{Research questions and hypothesis}

The overarching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.

\noindent
We develop the subquestions and their corresponding hypothesis separately in more detail.

@@

\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though after recording a SES gap increase of about 40\% in 30 years. \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap. First, we will concentrate on the evolution of the gap only for 15 year olds (which as we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap), which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}. Secondly, we will compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature. \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}. We posit no specific hypothesis for this question given that it's purely explanatory.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Which countries are experiencing changes in the achievement gap?

\end{enumerate}
\end{itemize}

\item The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because the bottom performers are falling behind or because both are changing at the same time. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. Following that strategy will put us closer to understanding the source of the evolution of the achievement gap.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Is the gap originating because the top is gaining ground, the top is falling behind or because of a dynamic interaction between the two?

\item Hypothesis 1: The theoretical argument in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups. We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-SES groups at a greater disadvantage. Conversely, in countries with low tracking we should expect for the achievement gap to be narrowing.

\end{enumerate}
\end{itemize}

\item Our study itself is composed of six cross-sectional surveys, which limits the design to single country snapshots, but given that most countries participated in all six surveys we can model the evolution of the achievement gap over time. We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking} but not empirically tested before in an international context.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Are tracking and vocational enrollment related to the evolution of the gap?

\item Hypothesis 2: Building on the previous hypothesis, tracking should play an important role on the evolution of the achievement gap. We hypothesize that the degree of tracking of an educational system is tightly related to changes in the gap, and the more tracking, the more inequality. Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.

\end{enumerate}
\end{itemize}

\item Lastly, we want to investigate whether better performing countries have lower levels of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high levels of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Are achievement and inequality correlated? and if so, has the relationship weakened/strengthen over time?

\item Hypothesis 3: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality. Countries with high performance should have less inequality and countries with low performance should have greater inequality

\end{enumerate}
\end{itemize}

\end{enumerate}

@@

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016. This paper takes advantage of these six waves to build a country pseudo-panel where each country-wave will have its corresponding achievement gap, making it possible to study changes in nearly 15 years for each country.

### 2017-10-10 paper/paper.tex
Rewrote hypothesis section + introduction

\title{Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\setlength{\parindent}{2em}
\setlength{\parskip}{1em}
\showboxdepth=5
\showboxbreadth=5

@@

% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}. But fast-forward until today, we've understood the relationship much better.

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it. For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavorable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\). That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns. Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment that yields the highest returns. This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities. We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

@@

\section{Research framework}

\subsection{Research questions}

This study seeks to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries and search for a likely explanation. This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems. In addition, it extends the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's an over-time tradeoff between good performance and equality over time separately for each country. The advantages of this study are threefold.

First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}. As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap. Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.

Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States. This will help assess the magnitude of the inequality as well as the rate of growth/decline. This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes. That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.

Third, we attempt to model why the achievement gap is changing over time. In contrast to other studies, this is a novelty given that most research has been done cross-sectionally between countries.

More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
d) What explains this cross-country evolution.

We develop each question separately for more detail.

a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap over time. We will compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}

b) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.

c) We want to investigate whether better performing countries have lower levels of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high levels of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

d) Tightly linked to the previous point, we want to test whether several dimensions of the tracking setup explain why the gaps are changing. This will support the previous points in suggesting that the dynamics of the over-time achievement gap are endogenously intertwined with the effect that tracking has on the specific changes of the high and low SES groups.

\subsection{Hypothesis}

Following the previous questions and the literature on educational inequality and tracking we pose three hypothesis for answering questions (b), (c) and (d) given that question (a) is purely an exploratory exercise.

\begin{itemize}
\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups. For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this does not mean that the top is lowering their score but that the bottom is most likely catching up).

\item Hypothesis 2: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality. Countries with high performance should have less inequality and countries with low performance should have greater inequality

\item Hypothesis 3: if any (or both) of the previous hypothesis are confirmed, then tracking should play an important role on the evolution of the achievement gap. We hypothesize that the degree of tracking is tightly related to changes in the gap, and the more tracking, the more inequality. Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.
\end{itemize}

@@

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

>>>>

\title{Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\setlength{\parindent}{2em}
\setlength{\parskip}{1em}
\showboxdepth=5
\showboxbreadth=5

@@

The impact of inequality, be it economic or of non pecuniary causes, is a crucial determinant for the opportunities of a child. Evidence from disciplines as disparate as epidemiology, psychology, sociology and economics keep piling evidence that suggests that inequality in the life course is something present in virtually all societies. This is something that politicians and policymakers are aware of and, yet despite their many efforts, inequality has not abated.

The literature has found conclusive evidence showing that, for example, the United States has increased both economic inequality \citep{piketty2003}, inequality of opportunity \citep{chetty2014} and inequality of achievement \citep{reardon2011}. All of these findings, however, are from the United States. Other countries such as the United Kingdom or other developed countries have their fair share of studies when there is data available. However, very rarely can we compare levels of inequality (of any kind), across several countries with a great degree of confidence.

This paper looks to study the cognitive achievement gap between the 90th and 10th percentile of the SES distribution. This gap can be termed as the SES achievement gap between the top and bottom SES groups. Using the PISA database, we calculate the achievement gap for every country and for all six available PISA waves, building a country panel that shows a time trend of 15 years for each country.

More concretely, we aim to study the evolution of the achievement gap. We aim to answer two broad question: Which countries are changing over time while others are stagnated? Equally important, what explains these changes? Building on the literature on educational institutions we suggest the degree of tracking of a country and it's level of vocational enrollments as possible explanations for the changes of a country over time.

Findings suggest that there's reasonable variation in the evolution of the gap. For example, the United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%. This reduction is also present in other countries such as Germany and Poland. Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap. Robustness checks shows that also the 80/20 and 70/30 replicate the results for nearly all countries.

We then assess whether this gap is correlated with the country's average performance, and find a very strong negative correlation for all country pooled. This suggests that countries with high levels of achievement have low levels of inequality. Yet these results are aggregated for every year. A more disaggregate analysis shows that this pattern is also present for every country over time, although some exceptions are encountered, like the United States where achievement and inequality as positively correlated.

To finish, we find that the degree of tracking of a country seems to be a likely explanation to the changes in the achievement gap, but it should not be studied without factoring in the degree of vocational enrollment. We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap. Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly. Simulations show that if every country switched to little or no tracking their gaps would shrink by about 20\%.

These findings have two specific implications. First, inequality should not be 'generalized' or 'summarized' with one single indicator for many countries. There are widespread differences between countries, where each one should be studied in context. Secondly, now that we know that, further research should attempt to explain why some countries follow such different evolutions. We propose a possible explanation, the degree of tracking and vocational enrollment of the country. Tracking seems to exacerbate inequality, but if coped with a reasonable level of vocational enrollment it can alleviate the negative effects of tracking. Further research should confirm these results hold for other empirical tests and if so, policymakers should incorporate vocational enrollment not only as a 'labor market' solution to youth unemployment but as a compensatory mechanism for highly tracked curriculums.

\section{Literature Review}
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}. But fast-forward until today, we've understood the relationship much better.

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it. For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. To explain these, and other findings, \citet{cunha2006} suggest that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\). Although this sounds logical, it's importance is not understood entirely. The model implies that investments on children are not perfect substitutes at different time points but rather complements. There are specific periods of skill formation where it is much more cost-effective to invest in, which are the early stages of the life course. In fact, when tested against the data, \citet{cunha2006} show that their explanation fits the results of most of the empirical literature very well.

For this reason, children who receive less (or worse) investments have a cumulative disadvantage against children who do receive quality investments. This leads, as early as preschool, to an unequal start in life. That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns \citep{elango15}. Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills. Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}. Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistence, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.

This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities. We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

@@

\section{Research questions and hypothesis}

The overarching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.

\noindent
We develop the subquestions and their corresponding hypothesis separately in more detail.

\begin{enumerate}

\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though after recording a SES gap increase of about 40\% in 30 years. \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap. First, we will concentrate on the evolution of the gap only for 15 year olds (which as we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap), which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}. Secondly, we will compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature. \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}. We posit no specific hypothesis for this question given that it's purely explanatory.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Which countries are experiencing changes in the achievement gap?

\end{enumerate}
\end{itemize}

\item The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because the bottom performers are falling behind or because both are changing at the same time. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. Following that strategy will put us closer to understanding the source of the evolution of the achievement gap.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Is the gap originating because the top is gaining ground, the top is falling behind or because of a dynamic interaction between the two?

\item Hypothesis 1: The theoretical argument in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups. We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-SES groups at a greater disadvantage. Conversely, in countries with low tracking we should expect for the achievement gap to be narrowing.

\end{enumerate}
\end{itemize}

\item Our study itself is composed of six cross-sectional surveys, which limits the design to single country snapshots, but given that most countries participated in all six surveys we can model the evolution of the achievement gap over time. We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking} but not empirically tested before in an international context.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Are tracking and vocational enrollment related to the evolution of the gap?

\item Hypothesis 2: Building on the previous hypothesis, tracking should play an important role on the evolution of the achievement gap. We hypothesize that the degree of tracking of an educational system is tightly related to changes in the gap, and the more tracking, the more inequality. Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.

\end{enumerate}
\end{itemize}

\item Lastly, we want to investigate whether better performing countries have lower levels of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high levels of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

\begin{itemize}
\item Research question:
\begin{enumerate}

\item Are achievement and inequality correlated? and if so, has the relationship weakened/strengthen over time?

\item Hypothesis 3: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality. Countries with high performance should have less inequality and countries with low performance should have greater inequality

\end{enumerate}
\end{itemize}

\end{enumerate}

@@

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016. This paper takes advantage of these six waves to build a country pseudo-panel where each country-wave will have its corresponding achievement gap, making it possible to study changes in nearly 15 years for each country.
